\pkg{DNest4}: Diffusive Nested Sampling in \proglang{C++} and \proglang{Python}

### 2016-05-28 paper/ms.tex
More veitch comments

%% an abstract and keywords
\Abstract{In probabilistic (Bayesian) inferences, we typically want to compute
properties of the posterior distribution, describing knowledge of
unknown quantities in the context of a particular dataset and the assumed
prior information. The marginal likelihood, also known as the ``evidence'',
is a key quantity in Bayesian model selection. The Diffusive Nested Sampling algorithm, a variant of Nested Sampling, is a powerful tool for generating
posterior samples and estimating marginal likelihoods. It is effective at
solving complex problems including many where the posterior distribution is
multimodal or has strong dependencies between variables.
\pkg{DNest4} is an open source (MIT licensed),
multi-threaded implementation of this algorithm in
modern \proglang{C++} (specifically, \proglang{C++11}),
along with associated utilities including: i)
a \pkg{Python} package allowing basic use without \proglang{C++} coding;
ii) \pkg{RJObject}, a class template for finite mixture models;
iii) Experimental support for models implemented in \proglang{Julia}; and
iv) An experimental \proglang{Python} tool to
generate \pkg{C++} code from a model specification.
In this paper we demonstrate \pkg{DNest4} usage through examples including
simple Bayesian data analysis, finite mixture models, and Approximate
Bayesian Computation.
}

@@

If prior information $I$ (hereby dropped)
implies a set of possible `models' $\{M_i\}$,
rather than a single one $M$, the posterior model probabilities are given by
\begin{align}
P(M_i | \data) &=
\frac{P(M_i)p(\data | M_i)}{\sum_j P(M_j)p(\data | M_j)}
\end{align}
where
\begin{align}
p(\data | M_j) &= \int p(\theta_j | M_j)p(\data | \theta_j, M_j) \, d\theta_j
\end{align}
is the marginal likelihood of model $j$. This kind of calculation is often
called ``model selection'' or ``model averaging''.

@@

A defining feature of NS is that it works with a sequence of
{\em constrained prior} distributions, proportional to $\pi$ but
restricted to regions of the parameter space where $L(\theta)$
is above some threshold $\ell$:
\begin{align}
p(\theta; \ell) &=
\frac{\pi(\theta)\mathds{1}\left[L(\theta) > \ell\right]}{X(\ell)}
\label{eqn:constrained_prior}
\end{align}
where
\begin{align}
X(\ell) &= \int \pi(\theta) \mathds{1}\left[L(\theta) > \ell\right] \, d\theta
\end{align}
is the amount of prior mass which has likelihood greater than $\ell$.
In the standard NS framework, the sequence of $\ell$ values is selected
so that $X(\ell)$ shrinks by a factor $\approx e^{-1/N}$ per iteration, where
$N$ is the number of particles used.

Sampling from the constrained priors (Equation~\ref{eqn:constrained_prior})
is required, and Markov Chain Monte Carlo (MCMC) is a popular method of doing
this, although alternatives exist \citep[e.g.][]{multinest, 2015MNRAS.453.4384H}.
Diffusive Nested Sampling \citep[DNS][]{dnest} is an alternative to NS for
problems where MCMC is the only viable sampling method. DNS is based on the
Metropolis algorithm. DNS evolves a point in the parameter space, along with
an integer index variable $j$,
to explore the following joint distribution:
\begin{align}
p(\theta, j) &= p(j)p(\theta | j)\\
&= w_j \times
\frac{\pi(\theta)\mathds{1}\left[L(\theta) > \ell_j\right]}{X(\ell_j)}.
\label{eqn:target_distribution}
\end{align}
where the $\{\ell_j\}$ are a sequence of increasing likelihood thresholds
or {\em levels}, and
$\ell_0 = 0$, and $\{w_j\}$ is the marginal distribution for $j$.
The marginal distribution for $\theta$ is then a {\em mixture} of
constrained priors:
\begin{align}
p(\theta) &=
\pi(\theta)\sum_{j=0}^{j_{\rm max}}
\frac{w_j\mathds{1}\left[L(\theta) > \ell_j\right]}{X(\ell_j)}.
\label{eqn:mixture_of_constrained_priors}
\end{align}
The DNS algorithm consists of two main stages. In the first stage,
the particle(s) are initialized from the prior $\pi(\theta)$, equivalent
to the mixture of constrained priors with a single level whose likelihood
threshold is $-\infty$. The mixture of constrained priors evolves by adding new levels
(see \citet{dnest} for details). In this stage,
the mixture weights $\{w_j\}$ are set according to
\begin{align}
w_j &\propto \exp(j/\lambda),\label{eqn:weighting}
\end{align}
where $\lambda$ is a scale length.
This enhances the probability that particles stay close to the
highest likelihood regions seen.
The second stage sets the mixture weights to be approximately
uniform ($w_j \propto 1$), with some tweaks described by \citet{dnest}.

@@

There are several ways of using \pkg{DNest4}. After installing the software
(Section~\ref{sec:installation}), you can implement a model as
a \proglang{C++} class
(Section~\ref{sec:models}) and compile it to create an executable file
to run \pkg{DNest4} on that problem. This method offers full control over the
design of your class, and allows the opportunity (for example) to optimise
performance by preventing a full re-computation of the log-likelihood when
only a small subset of the parameters were changed.

>>>>

%% an abstract and keywords
\Abstract{In probabilistic (Bayesian) inferences, we typically want to compute
properties of the posterior distribution, describing knowledge of
unknown quantities in the context of a particular dataset and the assumed
prior information. The marginal likelihood, also known as the ``evidence'',
is a key quantity in Bayesian model selection. The Diffusive Nested Sampling algorithm, a variant of Nested Sampling, is a powerful tool for generating
posterior samples and estimating marginal likelihoods. It is effective at
solving complex problems including many where the posterior distribution is
multimodal or has strong dependencies between variables.
\pkg{DNest4} is an open source (MIT licensed),
multi-threaded implementation of this algorithm in
\proglang{C++11},
along with associated utilities including: i)
a \pkg{Python} package allowing basic use without \proglang{C++} coding;
ii) \pkg{RJObject}, a class template for finite mixture models;
iii) Experimental support for models implemented in \proglang{Julia}; and
iv) An experimental \proglang{Python} tool to
generate \pkg{C++} code from a model specification.
In this paper we demonstrate \pkg{DNest4} usage through examples including
simple Bayesian data analysis, finite mixture models, and Approximate
Bayesian Computation.
}

@@

If prior information $I$ (dropped hereafter)
implies a set of possible `models' $\{M_i\}$,
rather than a single one $M$, the posterior model probabilities are given by
\begin{align}
P(M_i | \data) &=
\frac{P(M_i)p(\data | M_i)}{\sum_j P(M_j)p(\data | M_j)}
\end{align}
where
\begin{align}
p(\data | M_j) &= \int p(\theta_j | M_j)p(\data | \theta_j, M_j) \, d\theta_j
\end{align}
is the marginal likelihood of model $j$, equal to the expected value of the
likelihood function with respect to the prior distribution.
This kind of calculation is often
called ``model selection'' or ``model averaging''.

@@

A defining feature of NS is that it works with a sequence of
{\em constrained prior} distributions, proportional to $\pi$ but
restricted to regions of the parameter space where $L(\theta)$
is above some threshold $\ell$:
\begin{align}
p(\theta; \ell) &=
\frac{\pi(\theta)\mathds{1}\left[L(\theta) > \ell\right]}{X(\ell)}
\label{eqn:constrained_prior}
\end{align}
where
\begin{align}
X(\ell) &= \int \pi(\theta) \mathds{1}\left[L(\theta) > \ell\right] \, d\theta
\end{align}
is the amount of prior mass which has likelihood greater than $\ell$, and
$\mathds{1}()$ is the indicator function which takes the value 1 if the
argument is true and 0 otherwise.
In the standard NS framework, the sequence of $\ell$ values is selected
so that $X(\ell)$ shrinks by a factor $\approx e^{-1/N}$ per iteration, where
$N$ is the number of particles used. This geometric compression of the
parameter space is a defining feature of NS.

Sampling from the constrained priors (Equation~\ref{eqn:constrained_prior})
is required, and Markov Chain Monte Carlo (MCMC) is a popular method of doing
this, although alternatives exist \citep[e.g.][]{multinest, 2015MNRAS.453.4384H}.
Diffusive Nested Sampling \citep[DNS][]{dnest} is an alternative to NS for
problems where MCMC is the only viable sampling method. DNS is based on the
Metropolis algorithm. DNS evolves one or more particles in the parameter space, along with
an integer index variable $j$ for each particle,
to explore the following joint distribution:
\begin{align}
p(\theta, j) &= p(j)p(\theta | j)\\
&= w_j \times
\frac{\pi(\theta)\mathds{1}\left[L(\theta) > \ell_j\right]}{X(\ell_j)}.
\label{eqn:target_distribution}
\end{align}
where the $\{\ell_j\}$ are a sequence of increasing likelihood thresholds
or {\em levels}, and
$\ell_0 = 0$, and $\{w_j\}$ is the marginal distribution for $j$.
The marginal distribution for $\theta$ is then a {\em mixture} of
constrained priors:
\begin{align}
p(\theta) &=
\pi(\theta)\sum_{j=0}^{j_{\rm max}}
\frac{w_j\mathds{1}\left[L(\theta) > \ell_j\right]}{X(\ell_j)}.
\label{eqn:mixture_of_constrained_priors}
\end{align}
The DNS algorithm consists of two main stages. In the first stage,
the particle(s) are initialized from the prior $\pi(\theta)$, equivalent
to the mixture of constrained priors with a single level whose likelihood
threshold is $-\infty$. The mixture of constrained priors evolves by adding new levels
(see \citet{dnest} for details). In this stage,
the mixture weights $\{w_j\}$ are set according to
\begin{align}
w_j &\propto \exp(j/\lambda),\label{eqn:weighting}
\end{align}
where $\lambda$ is a scale length.
This enhances the probability that particles stay close to the
highest likelihood regions seen.
The second stage sets the mixture weights to be approximately
uniform ($w_j \propto 1$), with some tweaks described by \citet{dnest}.

@@

There are several ways of using \pkg{DNest4}. After installing the software
(Section~\ref{sec:installation}), you can implement a model as
a \proglang{C++} class
(Section~\ref{sec:models}) and compile it to create an executable file
to run \pkg{DNest4} on that problem. This method offers full control over the
design of your class, and allows the opportunity of optimizing
performance by preventing a full re-computation of the log-likelihood when
only a small subset of the parameters have been changed.
